---
title: "Analysis of Severe Weather Events for Human and Economic Impact"
output:
  html_document:
    keep_md: true
---
```{r setoptions,echo = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, restult = "hide")
```

##Synopsis: 
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

##Data Processing: 
The NOAA Storm Event Database (herein, database) records 37 variables on storm events from 1950 to present day.  This project will use [a subset of the data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) with nearly 1 million storm events from April 1950 to November 2011.  In addition to the dataset [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [FAQs](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf) are available.

###Start by downloading and reading in the database file.

```{r cache = TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
tempFile <- tempfile()
if(capabilities("libcurl")) {
        download.file(fileURL, tempFile, method = "libcurl")        
} else {
        download.file(fileURL, tempFile)
}
```

- To cut down on processing and cleaning time we will import only columns relating to the type of event, fatalities, injuries, and property damage.
```{r cache=TRUE}
#Grab the headers list.
headers <- read.csv(tempFile, nrows = 1)
#Ignore all columns except the ones we specify.
headers[,] <- c(rep("NULL", ncol(headers)))
headers[c("FATALITIES", "INJURIES", "PROPDMG", "CROPDMG")] <- "numeric"
headers[c("BGN_DATE", "PROPDMGEXP", "CROPDMGEXP")] <- "factor"
headers[c("EVTYPE")] <- "character"
#Read in the specified columns
StormDF <- read.table(tempFile, header = TRUE, sep = ",", quote = "\"", fill = TRUE, comment.char = "", colClasses = headers)
unlink(tempFile)

```

###Clean and preprocess the data to look for anomalies and obvious errors.
- Assign better column names.
```{r}
names(StormDF)[names(StormDF) == "EVTYPE"] <- "Event.Type"
names(StormDF)[names(StormDF) == "BGN_DATE"] <- "Begin.Date"
names(StormDF)[names(StormDF) == "FATALITIES"] <- "Fatalities"
names(StormDF)[names(StormDF) == "INJURIES"] <- "Injuries"
names(StormDF)[names(StormDF) == "PROPDMG"] <- "Property.Damage"
names(StormDF)[names(StormDF) == "PROPDMGEXP"] <- "Property.Damage.Exponent"
names(StormDF)[names(StormDF) == "CROPDMG"] <- "Crop.Damage"
names(StormDF)[names(StormDF) == "CROPDMGEXP"] <- "Crop.Damage.Exponent"
```

- Drop all rows with no Health or Economic loss
```{r}
StormDF <- StormDF[!rowSums(StormDF[c("Fatalities","Injuries","Property.Damage","Crop.Damage")])==0,]
```

- Check if there are missing values.
```{r}
anyNA(StormDF)
```

- Are the Property Damage and Crop Damage exponent fields accurate?
```{r}
unique(StormDF$Property.Damage.Exponent)
unique(StormDF$Crop.Damage.Exponent)
```
-- Does the number of incorrect values in "exponent" fields represent a statistically relevant set?
```{r}
allowedExp <- c("k", "K", "M", "m", "B", "b", "0", "0.0", "0.00", NA, "")
percErrors <- round(nrow(subset(StormDF, !Property.Damage.Exponent %in% allowedExp | !Crop.Damage.Exponent %in% allowedExp)) / nrow(StormDF), 5)*100

```

Events with erroneous exponent fields represent `r percErrors` percent of the data.

- Property and crop damage are reported with dollar amount and exponential units in separate columns.  Calculate the damage cost by multiplying the exponents and combining.
```{r cache=TRUE}
#Calculate the full property damage cost by multiplying the exponent
StormDF["Property.Damage.FULL"] <- StormDF["Property.Damage"]
StormDF[StormDF$Property.Damage.Exponent %in% c("k","K"),"Property.Damage.FULL"] <- 
        StormDF[StormDF$Property.Damage.Exponent %in% c("k","K"),"Property.Damage"]*1000
StormDF[StormDF$Property.Damage.Exponent %in% c("m","M"),"Property.Damage.FULL"] <- 
        StormDF[StormDF$Property.Damage.Exponent %in% c("m","M"),"Property.Damage"]*1000000
StormDF[StormDF$Property.Damage.Exponent %in% c("b","B"),"Property.Damage.FULL"] <- 
        StormDF[StormDF$Property.Damage.Exponent %in% c("b","B"),"Property.Damage"]*1000000000
#Do the same for crop damage
StormDF["Crop.Damage.FULL"] <- StormDF["Crop.Damage"]
StormDF[StormDF$Crop.Damage.Exponent %in% c("k","K"),"Crop.Damage.FULL"] <- 
        StormDF[StormDF$Crop.Damage.Exponent %in% c("k","K"),"Crop.Damage"]*1000
StormDF[StormDF$Crop.Damage.Exponent %in% c("m","M"),"Crop.Damage.FULL"] <- 
        StormDF[StormDF$Crop.Damage.Exponent %in% c("m","M"),"Crop.Damage"]*1000000
StormDF[StormDF$Crop.Damage.Exponent %in% c("b","B"),"Crop.Damage.FULL"] <- 
        StormDF[StormDF$Crop.Damage.Exponent %in% c("b","B"),"Crop.Damage"]*1000000000
```

-- Add columns to summarize economic and personal health impact.
```{r}
StormDF["Economic.Impact"] <- StormDF$Property.Damage.FULL + StormDF$Crop.Damage.FULL
StormDF["Health.Impact"] <- StormDF$Fatalities + StormDF$Injuries
```

- The dataset has three distinct collection periods.  
1. From 1950 to 1954 only Tornado events were collected.  
2. From 1955 through 1995, only tornado, thunderstorm wind and hail events were collected. From 
3. From 1996 to present (2011), 48 event types are recorded as defined in NWS Directive 10-1605 (http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf) were collected.

-- Reassign the Date column to allow for subsetting according to these collection periods.  We don't need to factor in the TIMEZONE column for the sake of subsetting at a rough level.
```{r}
StormDF$Begin.Date <- as.Date(StormDF$Begin.Date, format = "%m/%d/%Y")

#Create factor variable for subsetting.
StormDF[which(StormDF$Begin.Date <= as.Date("1954-12-31")),"Reporting.Period"] <- "1950 to 1955"
StormDF[which(StormDF$Begin.Date > as.Date("1954-12-31") & StormDF$Begin.Date <= as.Date("1995-12-31")),"Reporting.Period"] <- "1955 to 1996"
StormDF[which(StormDF$Begin.Date > as.Date("1995-12-31")),"Reporting.Period"] <- "1996 to Present"
StormDF$Reporting.Period <- as.factor(StormDF$Reporting.Period)
```

- Check the validity and consistency of the "Event Type" column.  According to the documentation there are 48 "correct" storm event types.
-- Create a list of allowed (known) Event Types from page 6 of http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf
```{r}
allowedEvents <- toupper(c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", 
                           "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", 
                           "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", 
                           "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", 
                           "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", 
                           "Waterspout", "Wildfire", "Winter Storm", "Winter Weather"))
```

- How many Event Types are listed in the dataset?
```{r}
StormDF$Event.Type <- toupper(StormDF$Event.Type)
percErrors <- round(nrow(subset(StormDF, !Event.Type %in% allowedEvents)) / nrow(StormDF), 3)*100
```
There are `r length(unique(StormDF$Event.Type))` unique "events" in the data.
Events with non-standard event fields represent `r percErrors`% of the data.

-- What is the magnitude of the economic and health impact for these "non-standard" events?
```{r}
skewEconomicImpact <- round(sum(subset(StormDF, !Event.Type %in% allowedEvents)$Economic.Impact) 
                            / sum(StormDF$Economic.Impact), 3)*100
skewHealthImpact <- round(sum(subset(StormDF, !Event.Type %in% allowedEvents)$Health.Impact) 
                         / sum(StormDF$Health.Impact), 3)*100
```
Events with non-standard event fields represent `r skewEconomicImpact`% of the total Economic Impact and `r skewHealthImpact`%  
of the total Health Impact of all storm events.


-- We need to clean up the egregious and obviously mis-labeled event types.  
NOTE: This needs to be done according to a specific order because of some events that have multiple event types in the name (ie. Thunderstorm / Hail).  The below order is based on assesment of the direct vs indirect effects of Economic and Health impact in decreasing order of magnitude.  
  
1. Primary cause: The major event ("Blizzard", "Funnel Cloud", "Hurricane (Typhoon)", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami",  "Wildfire", "Winter Storm"") are the causal events for secondary events such as wind, rain, flood, etc.  Also marine events are a special subset to start with.
```{r}
StormDF[grep("WATERS", StormDF$Event.Type),"Event.Type"] <- "WATERSPOUT"
StormDF[grepl("^\\s*TORNA", StormDF$Event.Type),"Event.Type"] <- "TORNADO"
StormDF[grep("(HUR|TYP)", StormDF$Event.Type),"Event.Type"] <- "HURRICANE (TYPHOON)"
StormDF[grep("TROPIC.*STORM", StormDF$Event.Type),"Event.Type"] <- "TROPICAL STORM"
StormDF[grep("BLIZ", StormDF$Event.Type),"Event.Type"] <- "BLIZZARD"
StormDF[grep("FUNNEL.*CLO", StormDF$Event.Type),"Event.Type"] <- "FUNNEL CLOUD"
StormDF[grep("TSU", StormDF$Event.Type),"Event.Type"] <- "TSUNAMI"
StormDF[grep("FIRE", StormDF$Event.Type),"Event.Type"] <- "WILDFIRE"
StormDF[grepl("WINTER", StormDF$Event.Type) & 
                !StormDF$Event.Type %in% c("WINTER WEATHER"),"Event.Type"] <- "WINTER STORM"
```
2. Secondary cause: A primary event may have multiple secondary causes for damage costal flood, flash flood, flood, hail, wind, storm surge, high surf, high winds, etc that infer a precedance on the impact that they have.
```{r}
#Cleanup the various flood types.
StormDF[grep("(COASTAL.*FLOOD|CSTL.*FLOOD|TID.*FLOOD)", StormDF$Event.Type),"Event.Type"] <- "COASTAL FLOOD"
StormDF[grep("BEACH.*FLOOD", StormDF$Event.Type),"Event.Type"] <- "COASTAL FLOOD"
StormDF[grep("(FLASH.*FLOOD|FLOOD.*FLASH)", StormDF$Event.Type),"Event.Type"] <- "FLASH FLOOD"
StormDF[grepl("LAKE.*FLOOD", StormDF$Event.Type),"Event.Type"] <- "LAKESHORE FLOOD"
StormDF[grepl("(FLOOD|FLD)", StormDF$Event.Type) & 
                !StormDF$Event.Type %in% c("FLASH FLOOD", "COASTAL FLOOD", "LAKESHORE FLOOD"),"Event.Type"] <- "FLOOD"

#Cleanup other outliers
StormDF[grepl("HAIL", StormDF$Event.Type) & 
                !StormDF$Event.Type %in% c("MARINE HAIL"),"Event.Type"] <- "HAIL"
StormDF[grep("HEAVY RAIN", StormDF$Event.Type),"Event.Type"] <- "HEAVY RAIN"
StormDF[grep("LAKE.*SNOW", StormDF$Event.Type),"Event.Type"] <- "LAKE-EFFECT SNOW"
StormDF[grep("HEAVY SNOW", StormDF$Event.Type),"Event.Type"] <- "HEAVY SNOW"
StormDF[grep("MAR.*TSTM", StormDF$Event.Type),"Event.Type"] <- "MARINE THUNDERSTORM WIND"
StormDF[grepl("(THUNDERST|^\\s*TSTM)", StormDF$Event.Type) & 
                !StormDF$Event.Type %in% c("MARINE THUNDERSTORM WIND"),"Event.Type"] <- "THUNDERSTORM WIND"
StormDF[grep("(EXT.*COLD|EXT.*WIND.*CHIL)", StormDF$Event.Type),"Event.Type"] <- "EXTREME COLD/WIND CHILL"
StormDF[grepl("WIND.*CHI", StormDF$Event.Type) & 
                !StormDF$Event.Type %in% c("EXTREME COLD/WIND CHILL"),"Event.Type"] <- "COLD/WIND CHILL"
StormDF[grep("STRONG.*WIND", StormDF$Event.Type),"Event.Type"] <- "STRONG WIND"
StormDF[grep("HIGH.*WIND", StormDF$Event.Type),"Event.Type"] <- "HIGH WIND"
StormDF[grep("ICE FOG", StormDF$Event.Type),"Event.Type"] <- "FREEZING FOG"
StormDF[grepl("FOG", StormDF$Event.Type) & 
                !StormDF$Event.Type %in% c("FREEZING FOG"),"Event.Type"] <- "DENSE FOG"
StormDF[grep("^SNOW$", StormDF$Event.Type),"Event.Type"] <- "HEAVY SNOW"
StormDF[grep("^STORM.*SURGE$", StormDF$Event.Type),"Event.Type"] <- "STORM SURGE/TIDE"
StormDF[grep("^FREEZING RAIN$", StormDF$Event.Type),"Event.Type"] <- "ICE STORM"
StormDF[grep("^WIND$", StormDF$Event.Type),"Event.Type"] <- "HIGH WIND"
StormDF[grep("^LANDSLIDE$", StormDF$Event.Type),"Event.Type"] <- "HEAVY RAIN"
StormDF[grep("RIP.*CUR", StormDF$Event.Type),"Event.Type"] <- "RIP CURRENT"
StormDF[grep("WARM", StormDF$Event.Type),"Event.Type"] <- "HEAT"
StormDF[grep("HEAVY SURF/HIGH SURF", StormDF$Event.Type),"Event.Type"] <- "HIGH SURF"
StormDF[grep("HIGH TIDE", StormDF$Event.Type),"Event.Type"] <- "STORM SURGE/TIDE"
```

-- After doing some cleanup what is the magnitude of mis-classified events?
```{r}
percErrors <- round(nrow(subset(StormDF, !Event.Type %in% allowedEvents)) / nrow(StormDF), 3)*100
```
There are now `r length(unique(StormDF$Event.Type))` unique "events" in the data.
Events with "non-standard" event fields represent `r percErrors` percent of the data.

  What is the magnitude of the economic and health impact for these "non-standard" events?
```{r}
skewEconomicImpact <- round(sum(subset(StormDF, !Event.Type %in% allowedEvents)$Economic.Impact) 
                            / sum(StormDF$Economic.Impact), 3)*100
skewHealthImpact <- round(sum(subset(StormDF, !Event.Type %in% allowedEvents)$Health.Impact) 
                         / sum(StormDF$Health.Impact), 3)*100
```
Events with "non-standard" event fields now represent `r skewEconomicImpact`% of the total Economic Impact  
and `r skewHealthImpact`% of the total Health Impact of all storm events.

- Since the first and second reporting period only captured a subset of the final 48 Event types we may want to treat each period as its own section.

Cross-tabulate the dataset based on fatalities, injuries, and total by reporting period.
```{r}
require(reshape2)
meltedDF <- melt(StormDF, id.vars = c("Event.Type", "Reporting.Period"),
                          measure.vars = c("Fatalities", "Injuries", "Health.Impact", "Property.Damage.FULL", "Crop.Damage.FULL", "Economic.Impact"))
summaryDF <- dcast(meltedDF, Event.Type + Reporting.Period ~ variable, sum)
```

- Though the improperly coded events do not represent a significant magnitude of the data set they do create lots of noice on the charts.  

Let's grab just the top N events for each reporting period for both Economic and Health impact.
```{r}
topN <- 5
topEconomic <- do.call(rbind, by(summaryDF,summaryDF$Reporting.Period, 
                                function(dat) dat[order(dat$Economic.Impact,decreasing=TRUE)[1:topN],]))
topHealth <- do.call(rbind, by(summaryDF,summaryDF$Reporting.Period, 
                               function(dat) dat[order(dat$Health.Impact,decreasing=TRUE)[1:topN],]))
```



##Results: 
```{r}
library(ggplot2)
library(grid)
topEconomicPlot <- ggplot(na.omit(topEconomic), aes(x = factor(Event.Type), y = Economic.Impact, fill = factor(Reporting.Period))) +
        geom_bar(stat = 'identity')  + 
        ylab('Economic Impact (USD)') + 
        xlab('') +
        theme(legend.position = 'none',
              plot.margin = unit(c(3, 1, 3, 1), "lines"),
              axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

topHealthPlot <- ggplot(na.omit(topHealth), aes(x = factor(Event.Type), y = Health.Impact, fill = factor(Reporting.Period))) +
        geom_bar(stat = 'identity')  + 
        ylab('Health Impact (Fatalities + Injuries)') + 
        xlab('') +
        theme(legend.position = 'bottom', 
              ##legend.text.align = 'vertical', 
              plot.margin = unit(c(3, 1, 1, 1), "lines"),
              axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
        scale_fill_discrete(guide = guide_legend(title = "Reporting Period"))

pushViewport(viewport(layout = grid.layout(1, 2)))
print(topEconomicPlot, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(topHealthPlot, vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
grid.text("Event Type", y = 1, just = c("center","bottom"))
grid.text("Top N Storm Event Types by Economic and Health Impact", y = 1, just = c("center","top"))

```
