---
title: "Evaluation and Simulation of the Exponential Distribution"
output:
  html_document:
    keep_md: true
---

```{r setoptions,echo = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, restult = "hide")
```

####by Michael Gregory

###Overview: In this project I will investigate the exponential distribution in R and compare it with the Central Limit Theorem.  I will compare the population and sample mean as well as the population and sample variance.

###Simulations: The exponential distribution will be simulated in R with rexp(n, lambda) where lambda is the rate parameter. 

-Using a lambda of .2, generate 1000 simulations of 40 random exponential values and also calculate the means and variances of those 1000 simulations.
```{r}
set.seed(11072015)

nosim <- 1000
n <- 40
lambda <- 0.2
sampleMatrix <- matrix(rexp(nosim * n, lambda), nosim)
sampleMeans <- apply(sampleMatrix, 1, mean)
sampleVars <- apply(sampleMatrix, 1, var)
```

###Sample Mean versus Theoretical Mean: 
-From the exponential distribution we know that the distribution mean is 1/lambda.
```{r}
distMean <- 1/lambda
distMean
```

-For the sample mean we know that:
$$
\bar X = \sum_{i=1}^n x_i p(x_i)
$$
where $p(x_i) = 1/n$

We could use the mean() function in R but lets calculate it using the function.
```{r}
calcSampleMean <- sum(sampleMatrix) / (nosim * n)
calcSampleMean
sampleMean <- mean(sampleMatrix)
sampleMean
meanDiff <- round((calcSampleMean - sampleMean), 10)
```

Rounded to 10 decimal places we confirm that the difference between our calculation of the mean and R's mean() function is `r meanDiff`.  

Likewise, we see that the sample mean (`r round(calcSampleMean,4)`) is a good approximation of the distribution mean (`r distMean`).  
  
-From a plot we can see that the distribution of the sample means appears to be centered around the distribution mean.
```{r}
par(mfrow = c(1,2))
hist(sampleMatrix, main = "Histogram of Exponential Variables", xlab = "Exponential Variables")
abline(v=distMean, lwd = 4, col = "red")
legend("topright", "Distribution Mu = 5", lty=1, lwd=2.5, col= "red")

hist(sampleMeans, main = "Histogram of Sample Means", xlab = "Sample Means")
abline(v=distMean, lwd = 4, col = "red")
```

###Sample Variance versus Theoretical Variance: 


-From the exponential distribution we know that the distribution SD (sigma) is 1/lambda and the distribution variance is sigma^2.  
```{r}
distSD <- 1/lambda
distVar <- distSD^2
distVar
```

-For the sample variance, we could use R to calculate the sample variance but lets do it manually first.
```{r}
calcSampleVar <- sum((sampleMatrix - sampleMean)^2) / ((nosim * n) - 1)
sampleVar <- var(c(sampleMatrix))
round((calcSampleVar - sampleVar), 10)
```
Here we see that the sample variance (`r round(sampleVar, 4)`) is a good approximation of the distrbution variance (`r round(distVar, 4)`).  

#Calculate the sample variances and graph them to see the distribution.

```{r}

sampleVar <- apply(sampleMatrix, 1, var)

```



###Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.

???Central Limit Theorem.  ???
Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

```{r}
par(mfrow = c(1,2))
hist(sampleMatrix, main = "Histogram of Probabilities of Exponential Variables", xlab = "Exponential Variables", prob = TRUE)
curve(dexp(x,lambda), add=TRUE, col = "blue")

hist(sampleMeans, main = "Histogram of Probabilities of Sample Means", xlab = "Sample Means", prob = TRUE)
curve(dnorm(x, mean = sampleMean), add=TRUE, col = "blue")
```